The Spine of American Legal Practice

### 2016-06-18 article/article.Rmd
Import from Google Docs

---
title: 'The “Paste-Pot” Field Code: The Migration of Law in the Nineteenth-Century
  United States'
author: "Kellen Funk and Lincoln Mullen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

>>>>

---
title: The Spine of American Legal Practice
author:
- Kellen Funk
- Lincoln Mullen
output: pdf_document
...

At the opening of the first Nevada legislature in 1861, Territorial Governor James W. Nye, a former New York lawyer, instructed the assembly that they would have to forsake the Mormon statutes of Utah Territory from which Nevada had been carved. Those laws were ill adapted to "the mining interests" of the new territory, but "happily for us, a neighboring State whose interests are similar to ours, has established a code of laws" sufficiently attractive to "capital from abroad." That neighbor, of course, was California, and Nye urged that California's code of civil practice be enacted in Nevada, as far as it could "be made applicable."[^1] Territorial Senator William Morris Stewart, the famed silver-mining lawyer who would become a leading U.S. Senator of Reconstruction, followed the instructions perhaps too well. Stewart literally cut and pasted the latest *Wood's Digest* of the California Practice Act into the session bill, crossing out *state* and *California* and substituting *territory* and *Nevada* where necessary. Stewart copied not just California's code but also its method of codification---California too had borrowed its code in a similar manner from New York.

Nye wrote back to the assembly in disgust. The bill---of 715 sections---had reached him late the night before the legislative session was to close. Even in the few hours he had to read it, Nye counted "so many errors in the enrolling of it, numbering probably more than three hundred." Some errors were severe. Nevada's Organic Act specified the jurisdiction of the district courts and justices of the peace, but the code overwrote these limitations by copying California's arrangements. Error-riddled and unconstitutional as the bill was, Nye believed a civil practice code was a "universal necessity and public need," concluding that "without some legislation in providing a code of procedure in civil cases, it may be a matter of grave question whether your courts would be able to fulfill the purpose of their creation."[^2] Nye signed the Code into law. Strikingly, nothing like this "universal necessity" had existed when Nye began his legal career in New York in the 1840s.

By the end of the nineteenth-century, thirty-one American states and territories had adapted the text of a New York Code of Civil Procedure first promulgated in 1848. The Code colloquially became known as the Field Code, named for its chief draftsman David Dudley Field, a famed Manhattan trial lawyer.[^3] When Field's Code appeared in the Colorado assembly in 1876, a Denver newspaper wryly commented that "the scissors and paste-pot we had heretofore confidently believed were implements peculiar to the newspaper sanctum." But to the editor, the cut-and-paste code was not just a humorous curiosity of legislative history. Rather, the extensive copying of Field's Code threatened the foundations of American popular sovereignty. "The bill is a long one; the assembly has not the time to devote to it and to give it the scrutiny that a measure of such importance demands," the editor explained. But blind approval "would be an injustice to themselves and a greater wrong and injustice to the people who have a right to demand that their public servants legislate for the public good." The Code "has been 'assimilated,' as we are informed, 'to the character and requirements of our people,' whatever that may mean," but the editor feared the legislation was the product of "men who have the welfare of the 'dear people' at their tongue's end always, but never in their hearts."[^4]

The migration of the Field Code was a central event in Anglo-American legal history, but no modern historian has traced the extensive borrowings of the Field text nor recognized the political furor that greeted the Code outside New York.[^5] Every aspect of a civil justice system, the rules granting access to courts, to lawyers, to remedies (whether damages, injunctions, or seizure of property) was covered by the Code, making its New York-specific rules politically contentious both in- and outside the Empire State. As the Field Code migrated around the country, commentators in each jurisdiction raised the same complaint: How could legislation borrowed from another state represent the popular will and bests interests of *this* state?

Exploring this history is a daunting task, however. Procedure codes were long, technical documents, and although each jurisdiction copied large swaths of text, each also modified the text along the way, sometimes with a simple *Nevada* for *California*, sometimes with more foundational changes to the remedial system. Although Stewart's cut-and-paste code found its way into the archives, most draft legislation did not, and few codifiers explained in detail how they produced their texts.

Innovative techniques in digital analysis of texts can resolve some of these difficulties. Within a corpus of legislation, digital programming can essentially reverse engineer and visualize what the archive revealed in Figure 1: which texts were borrowed, which modified, and how extensively. Our method works especially well for legal texts, for reasons we will explain. However, we designed a replicable method that can offer similar results on any other kind of curated collection of texts, from church hymns to medical treatises.

The first contribution of this Article is the demonstration of our methods, applied to the Field Code. The second contribution is the integration of these methods with more conventional approaches of social and cultural history. On the national level the extent of legislative borrowing followed a pattern American historians have described as "Greater Reconstruction"---the former Confederate South and the Far West show a remarkable kinship. Scholars have typically described Greater Reconstruction as a *federal* development: the creation of national citizenship, a national economy, and a larger state apparatus centered in Washington, D.C. This Article shows that Greater Reconstruction had its state-level dimensions as well. The uniform practice of law and adjudication of civil remedies was not structured by Washington mandates, however, but by the anxiety that New York capital would follow only New York civil remedies. At the more local level, our digital computations can trace modifications within code traditions, for instance, the ways western and midwestern codifiers altered New York's law to accommodate hardening conceptions of racial competencies in the civil courts. To see these developments, one needs a tool that can canvass all the states at once across broad swaths of civil law.

**Part 1.**

Until 1848, civil remedies and trial practice in New York were largely governed by common law traditions loosely categorized as "practice and pleadings."[^6] To understand how to file a civil claim or to enforce a judgment, a lawyer had to consult ad hoc statutes from colonial times to the present as well as precedents reported from cases litigated at common law and chancery. By the 1840s, enterprising practitioners had collated these materials into a half dozen marketable treatises, but these remained works of private opinion---no court was bound to agree with the treatise writers as to the weight, relevance, or proper interpretation of a legal statement.[^7] The common law was accordingly known as "unwritten law" despite the proliferation of published texts. The point was that common law was not precisely determined until a particular case drawn from the messy facts of life demanded resolution.[^8] Statutes, on the other hand, were "written law," prescribing or reforming the rules even before a case put the precise question in issue. And within the realm of written law, codes were the ultimate statutes.[^9]

Codification is, as Lawrence Friedman has written, "one of the set pieces of American legal history." Law reformers advocated for a codification of the common law from the earliest days of the Republic up into the Gilded Age, from Massachusetts down to South Carolina. Efforts ranged from mere compilations of existing statutes in each state to a full European-style codification meant to be an entirely comprehensive and systematic statement of the law.[^10] At its most basic level, codification proposed that legislative policy ought to be the sole source of law. Law was to be made by democratically responsible legislators---in terse, unambiguous statements---not discovered through application and analogy in particular cases by the judges. Codification debates thus ranged over the metaphysics of law to the political theories of separation of powers and institutional competency.[^11]

At New York's constitutional convention in 1846, "the conquerors took all," as the ambivalent Jacksonian James Fenimore Cooper complained.[^12] Law reformers abolished the court of chancery, made judges to stand for popular election, and required the legislature to appoint a commission to reform the "practice and pleadings" of the civil courts.[^13] New Yorkers had two models of legislative commissions on which they could draw for their own law reforms. The French government under Napoleon had appointed five-member commissions to codify the law of France. When the New York law reformer William Sampson called for codification in a widely noted address to the New York historical society, one of the members of the French commission was living in exile in upstate New York at the time and wrote to Sampson to advise him on the mechanics of codification:

&gt;&gt;Courage; persevere in the support of written reason against precedents and vague traditions. . . . Let four or five good heads be united in a commission, to frame in silence the project of a code. It is not so difficult a task. It is only to consult together, and to select. Do so with your best authors as we did with ours, and principally with Pothier's treatise on Obligations, which we simply converted into articles of our code.&lt;&lt;

Tellingly, the French commissioner took it as granted that the commission's code would automatically be promulgated as law.[^14]

The other model came from England. Royal commissions had been employed since before the Revolution of 1688 to advise on a variety of matters. After the Revolution, royal commissions became disfavored while Parliament undertook many of the same functions through "committees of enquiry" to which only members of Parliament were appointed. By the nineteenth century, the use of royal commissions was again reviving. Royal commissioners were usually not legislators. On the one hand, that meant commissioners were not bound to the rhythms of Parliament's schedule---they could continue their work even as Parliament adjourned. On the other hand, the recommendations of royal commissioners were a further remove from actual legislation. Although by the mid-nineteenth century royal commissions sometimes offered model statutes, their advised legislation still had to pass through the normal politicking and drafting processes of Parliament.[^15]

After David Dudley Field and two other lawyers were appointed to the procedural commission, their reports made clear that they favored the French model but understood political realities would hold them to the English model. From 1847 to 1850, the commissioners made five reports to the legislature knowing they had no power to keep legislators from amending their code or even defeating it altogether. Each time they reminded legislators that "public opinion had issued its mandate in the most imposing form" of a constitutional decree. The constitution, together with the legislative act appointing the commission "gave the commissioners instructions so precise, as to leave them no discretion, if they had desired it, \[and\] promised them therefore in advance, so long as they obeyed those instructions, the concurrence and co-operation of all departments of the government."[^16] Although the theory of codification made it a democratic enterprise, in practice Field---an ardent Jacksonian---insisted the democratic legislature ought to defer to the expertise of the commissioners.

The Field commission sought to blunt criticism by insisting that political concerns about lawmaking did not apply to mere *procedure*. "The system of procedure by which law is administered, differs from the law itself in this," the commissioners explained: "the latter is a body of elementary rules founded in the immutable principles of justice, drawing their origin from the obligations which divine wisdom has imposed . . . ; while the former consists, in its very nature, but of a body of prescribed rules, having no source but the will of those by whom they are laid down." Substantive law was universal, natural, grounded in divine justice, and therefore entitled to respect and protection from change. But God cared nothing of the "the mere machinery by which law is to be administered," and thus, the commissioners argued, procedure was trivial enough for legislative experimentation but complicated enough that only master practitioners like themselves could run the experiment.[^17]

The scope of the Code tended to belie the idea that procedure included only the "mere machinery" of a lawsuit.[^18] The final draft of the Code, printed in 1850, spanned nearly 800 pages of 1,885 regulations. The first third of the Code covered constitutional topics, specifying the jurisdiction of all state courts and the duties of all state officers (and liabilities for violating those duties). The Code deregulated attorney compensation, introducing novel structures of retainers and contingency fees.[^19] It created summary procedures meant to accelerate debt collections while simultaneously carving out "homestead" exemptions from the sheriff's reach.[^20] The Code concluded by defining who could be an attorney, a juror, and a witness, drawing racial and gendered distinctions over who could speak in court.[^21]

Most importantly, the Code defined all the remedies that a civil court could order---from money damages, to partition of property, to injunctive decrees and contempts--and made those remedies available in every lawsuit. In many cases, a legal right was indistinguishable from the remedy that secured that right: The right to possess a particular piece of property and the remedy that seized and delivered that property were, in effect, the same thing. Remedies were thus intimately connected to substantive law. For that reason, neither the French *code de procédure civile* (1806) nor Blackstone's or Bentham's writings conceived of remedies as purely procedural.[^22] By codifying remedies, Field invited continual expansion of the category of "procedure." As other states adopted the procedure code, they sometimes included other fields of law that seemed paradigmatically "substantive" yet had such specific procedures or remedies that they too were placed in a "code of procedure," including the law of wills, corporations, or mortgages.[^23] And after all, argued procedural codifiers in Iowa, what did the famed Married Women's Property Acts offer besides procedural reform? These acts gave women *standing* to litigate in their own name and seek *remedies* in claims of property and contract, and they abolished mandatory rules of *joinder* (of husbands). Thus, one of the most significant changes to the law of property and domestic relations in the century went into the state's Code of Civil Procedure.[^24]

Despite Field's arguments that a mere procedure code was democratically unproblematic, his efforts were not entirely successful in New York. The commission submitted a draft of its main reforms in 1848, emphasizing that this first code was "but a report in part." New York's legislators enacted the partial code with little amendment, some legislators repeating Field's view that the constitution obligated them to accept the code.[^25] But when the commissioners submitted their final draft, the Assembly judiciary committee balked, directly disputing the commissioners' claims that procedure was merely the machinery of the law. They wrote that "provisions for rights and for the mode of pursuing remedies, insensibly run into each other," and distinctions between substance and procedure were difficult to draw in practice. In fact, "the parts of a system of law are more dependent on each other, and run into and affect each other, infinitely more than any machine of human contrivance." Procedure, at least in the common law tradition, was no mere machinery---it involved the entire legal apparatus. They therefore suspected the commissioners' forthcoming code of criminal procedure would include all of the criminal law as well, "as they seem to understand practice and pleadings to include all the law upon a given subject." That being the case, the committee wondered whether, "responsible as the Commissioners undoubtedly are, should we place in them a blind and implicit confidence that shall commit to their discretion the peace and property, the personal liberty and the lives of those who sent *us* here to make laws for them?"[^26]

The Code would encounter similar difficulties in each jurisdiction that adopted it. Even the shortest version of the Field Code was significantly longer than any other state statute before the Progressive legislation of the twentieth century. And unlike statutory complilations that sometimes took the name of a "code" but made no changes to existing law, the Field Code opened with a declaration of novelty, abolishing the hallmarks of prior practice and instituting "hereafter" a new form of action with substantial revisions to basic matters of standing, remedy, and enforcement of civil judgments.[^27] In the states where it was imported, there could be no getting around the fact that the Code introduced extensive, new law, yet legislators were unable to read, critique, and amend the Code within the short time horizons of a legislative session. "It is folly to undertake to pass a code in a sixty day session," wrote the *Montana Post*, "and the best way would be for the Assembly to select one from a State or Territory which would come near meeting our wants, and slide it through with the fewest changes possible."[^28] Sliding the Code through eased the problem of time but exacerbated the problem of local sovereignty. "To be governed by a foreign law, especially when that law is not preknown to the people whose conduct is to be regulated thereby . . . is something repugnant to the idea of Democratic Republican government," complained the *The Miner's Express*.[^29]

How, then, did states and territories achieve a politically acceptable balance between efficiency and sovereignty, borrowing law for sake of time but endowing it with popular legitimacy in each locale? Quite apart from the technicalities of legal practice, answering that question historically is daunting. The American federation of civil government into (depending on the period) more than thirty or forty separate jurisdictions creates significant difficulties for describing a phenomenon that was truly national despite its state-centered enactments. It requires some sense of how much law was borrowed in each locality and to what degree and significance innovations were introduced. But precisely because these questions concern codes---texts that comprehensively and systematically cover a given subject---they are ideal for innovative techniques in digital history.

**Part 2.**

Although the amount of legislation that governed American civil practice is impressive---even if uncodified, every state had procedure statutes that were amended and re-enacted nearly every decade---a corpus of procedural legislation would not even count among digital historians as "big data." Our collection of relevant (or possibly relevant) laws comprise 135 statutes from the nineteenth century. It includes separately bound codes of civil procedure as well as codes or regulations appearing session laws and statutory compilations.[^30] It comprises some **\[NUMBER\]** pages, **\[NUMBER\]** words, organized into 98,000 regulations. The corpus is not exhaustive. It includes the initial enactment of every code of civil procedure, as well as a sampling of procedure statutes and re-enacted codes from states reputed to have been legally influential, but it does not include every known statute of American procedural law. While a comprehensive project may be illuminating in its own ways, our specific question of how New York legislation influenced other states permits a more curated corpus.

Curating a corpus to answer a question is an important difference from the way that digital history work often describes itself. Speaking in broad terms, digital historical work---like all historical work---can either begin with a set of sources or a question.[^31] Digital history often emphasizes that it can begin with large datasets---the so-called "big data" which recently achieved the level of buzzword---such as the Google Books or Hathi Trust corpora.[^32] We think that this is indeed a worthwhile endeavor, and salutary for historians to have their questions shaped by the sources.[^33] But it is less apparent that historians can apply digital history methods to questions that they have already formed, or that the only thing that these methods, most notably of "distant reading" can do is to work at the absolute largest scales. Our method, which might unimaginatively named "medium data" is a useful way for historians to proceed, by finding a corpus which is bounded by a particular problem.

Because most codes were public statutes, they were widely printed and distributed and therefore found their way into libraries digitized by Google Books. Where possible, we collected our corpus from Google's publicly accessible database. We used optical character recognition software (OCR) to create plain-text versions of the codes, which we edited lightly---we corrected section markers by hand as necessary and wrote a script to fix the most common OCR errors.[^34]

The most important step we took in processing the files was to split each section of the code into its own text file. Codes varied in their organizations, but they universally divided specific regulations into *sections* (or, on occasion, *articles*). The discursive formation of these texts not only provides a handy organizational scheme for digital approaches, sections were historically the way legislators borrowed their texts.[^35] Codifiers took their sources apart by sections, rearranging here, editing, drafting, and then re-combining there. Despite the fact that states differed widely on what topics they included in "civil procedure," sectioning the codes allowed us to assess similarity even among codes of quite different lengths and coverage. For instance, we know from historical evidence that California's 1851 code was derived from New York's 1850 code. (Stephen Johnson Field, David Dudley's brother, was the lawyer who imported New York's code into California.[^36]) But the New York code is over 150,000 words long, where was California's code was just over 50,000 words long. Those rather disparate lengths mean that comparing all of the California code to all of the New York code is a less meaningful comparison that than comparing each section in the California code to each section in the New York code, where matching sections will be a similar length.

The next step was to compare sections to one another and measure the similarity between them. To continue the New York-to-California example, consider the following pairs of sections. The first pair is from the final draft of the New York Field Code. These sections completely abolish prior practice and begin to rebuild the procedure system from the ground up:

&gt;&gt;§ 554. The distinction between actions at law and suits in equity, and the forms of all such actions and suits, heretofore existing, are abolished; and, there shall be in this state, hereafter, but one form of civil action, for the enforcement or protection of private rights and the redress or prevention of private wrongs, which shall be denominated a civil action. § 555. In such action, the party complaining is known as the plaintiff, and the adverse party as the defendant.&lt;&lt;

In the theory of Euro-American lawyers, California had no prior practice to abolish, so the code began more simply:

&gt;&gt;§ 1. There shall be in this State but one form of Civil Action, for the enforcement or protection of private rights, and the redress or prevention of private wrongs. §2. In such action the party complaining shall be known as the plaintiff, and the adverse party as the defendant.&lt;&lt;

The pairs are obviously related to one another, both in terms of their legal force and in terms of the actual words used.[^37]

A well-known method for measuring the similarity of two documents involves dividing texts up into "tokens" of consecutive words and calculating a Jaccard score, defined as ratio between the number of tokens that the two document have in common to the total number of tokens that appear in both documents. We used five-word tokens and "shingled" them, meaning that for the New York sections above, the first token was "the distinction between actions at," the second token was "distinction between actions at law," and so on. These tokens each contain more meaning than a single word, yet because they are shingled they are robust to changes in the text or noisy OCR. A Jaccard score (in set notation, EQUATION) will always produce a number between 0 (complete dissimilarity) and 1 (complete similarity).

Comparing the section pairs above produces the following matrix of scores.

| NY1850-554 | NY1850-555 | CA1851-001 | CA1851-002 |

|:----------|:----------:|:----------:|:----------:|:----------:|

|NY1850-554 | | 0 | 0.14 | 0 |

|NY1850-555 | | | 0 | 0.41 |

|CA1851-001 | | | | 0 |

|CA1851-002 | | | | |

As we would expect the first respective sections (New York § 554 and California § 1) have a score of 0.14, which indicates that they are similar but have significant differences, while the second sections (New York § 555 to California § 2) have a much higher similarity score since only a few words were changed. Just as important when we compare the first section in New York to the second section in California, we get a score of 0; the two sections are nothing like each other.

The aim, then, was to create a triangular matrix like the one above, but with approximately 98,000 rows and 98,000 columns, containing the similarity scores for each possible pair of sections. While this is easy enough to conceptualize, such a matrix is actually quite large, containing about 4.8 billion comparisons. This would take an unreasonable amount of computation time, and most of these comparisons would be unnecessary, since each section has no relationship to most other sections (like the matrix above, most scores will be zero). We therefore implemented the minhash/locality sensitive hashing (LSH) algorithm to detect candidate pairs, that is, documents which were likely to be matches, and then we measured the Jaccard score for only those candidate pairs.[^38]

The result was a matrix of matches, with over 45,000 genuine matches. For each section in the corpus, after making some adjustments to remove anachronisms and spurious matches,[^39] we were able to identify the section from which it was most likely borrowed. In other words, we had traced the work of the commissioners' scissors and paste-pots through the course of their codes.

**Part 3.**

Having developed a method for finding legislative borrowings, we next sought to visualize the broader patterns revealed by the those borrowings. We did this first by aggregating the section to section borrowings into code to code borrowings. By counting how many sections each code borrowed from each other code, we were able to show the connections from one code to another.

The resulting network graph (**Figure X: Network Graph**) tells the genealogy of civil procedure in the United States. The New York codes, especially the finished draft of 1850, are central to the entire network.[^40] Yet it is important to note that New York spurred different traditions within the procedural network. Variations in the Field drafts meant that different states could borrow different versions of the Field Code. The 1850 New York code was the primary progenitor of several families of codes in California, Kentucky, Iowa, and Ohio, each of which in turn became major contributors to the law of neighboring states. The 1851 New York code became the progenitor of codes for Wisconsin, Florida, North Carolina, and South Carolina. Mississippi and Missouri borrowed from the New York 1849 code. These variations point out how the Field Code was always a moving target. While Field considered the 1850 New York code to be the definitive, ideal version of the code, it was really all of the New York codes from 1848 to 1853 that became models for other jurisdictions, likely due to whatever material a given commission had on hand. The Field Code was not a single volume on the shelf, but a series of volumes any one of which might be more accessible in different regions and different eras.

Even later New York codes can be considered a separate family. In 1876 a New York commission produced a new code attempting to consolidate all the statutory amendments subsequent to the 1851 Field Code. David Dudley Field was upset by the changes introduced in this revised code, called the Throop Code. A count by a "friend" of Field's found that only three sentences of the Field Code had carried over word-for-word into the Throop Code. Nevertheless, many of the changes were purely formal, and Field speculated that these changes had been made only "so that \[the Code\] should not appear to be the same thing as before." With respect to Mr. Field or his "friend," we were able to find that the connection was between the Throop Code and the Field Code was somewhat stronger than he thought, and the Throop Code was indeed a descendant, if much revised, from the Field Code.

Finally, our corpus included a number of statutes which stood outside the Field Code tradition, such as Virginia and West Virginia regulations, statutes from Massachusetts and Maine, and southern codes from Georgia to Louisiana. These statutes show that the dominance of the Field Code was not total, and a number of older jurisdictions remained outside of its ambit.[^41] But virtually every jurisdiction established or reconstructed after 1850 became a part of the the Field Code network, and no other tradition achieved anywhere near the same coherence across state lines.

We have so far discussed code to code borrowings which give us a macroscopic view of the corpus as a whole. But because the fundamental unit of comparison is section to section, it is possible to get a more detailed look at the kinds of borrowings that codes made of one another, by looking at each section in a code and seeing which other code it was borrowed from. These visualizations show how each code was put together. Codes at different removes from New York tended to have different patterns of borrowing. To illustrate this, we will follow one branch of the Field Code network, beginning with the family started by California's 1851 code.

<!-- ![](media/image04.png){width="6.5in" height="3.7916666666666665in"} -->
California's 1851 code borrowed heavily from two sources. The bulk of the code was taken directly from New York's 1849 or 1850 codes (which are difficult to distinguish exactly from one another.[^42] The remainder of the code was borrowed from California's 1850 code. The 1850 version was in turn borrowed from the New York codes \[**might as well show this, esp. since CA1850 is so impressive**\]. As we would expect, the bulk of the borrowings that come from California instead of directly from New York are at the beginning of the code, where the 1850 code had made amendments specific to the borrowing from California. (Many of the non-matching sections are tables of contents.) The main point is that that the majority of the code was lifted more or less directly from New York, with minor amendments specific to California. California made few to no innovations to the Field Code beyond a rearrangement of its provisions.

<!-- ![](media/image05.png){width="6.5in" height="4.291666666666667in"} -->
The pattern of borrowings in the Washington 1855 code was a rather different case. The Washington code was definitely in the lineage of the 1851 California code, since it borrowed sections from both California directly and from Oregon, which was derived from California. While there are a number of "other" sections, the pattern of borrowing is clear: Indiana's 1852 code and Oregon's 1854 code provide the majority of the borrowings. Those two contiguous bands of borrowings correspond to regulations on judgement from Oregon and enforcement provisions from Indiana. This pattern likely came about because one of the Washington code commissioners, Edward Lander, was an Indiana appellate judge from 1850 to 1853, while another commissioner, William Strong, was a justice of the Oregon Supreme Court in the same years. While working on the Washington code, they must have each used the law (and the law books) that they knew best. As a second generation variation on a regional code, the Washington code drew from different sources.

Finally, we can consider the case of Washington's revised 1873 code as an example of one of the last "leaves" on the family tree of the Field Code. The bulk of this code was lifted from the Washington 1855 code with only small amendments. The main exception was a probate code drawn from California's 1872 code.[^43] Like many of the last generation codes, the text of the procedure code had stabilized in a local manifestation of a regional tradition. The code was still genuinely a Field Code, with a great deal of similarity to the original New York Field Codes, but its specific form depended on the many edits and rearrangements that state code commissioners had made to the text.

<!-- ![](media/image03.png){width="6.5in" height="4.666666666666667in"} -->
So far we have retained the context of the surrounding sections within a particular code. But since our fundamental unit of comparison is section to section, we can use a technique called clustering to deform the texts (as a literary scholar might say) in order to see the textual history of how particular regulations embedded in individual sections changed over time.[^44]

We used a clustering algorithm to group related sections together. Clustering is the process of taking a set of objects (in this case, sections of the field codes) which have distances between one another (in this case, the Jaccard similarity is easily converted into a distance measure), and grouping items which are are most alike together. There are innumerable clustering algorithms, but we needed one that could work with a sparse matrix and one whose assumptions matched the characteristics of our problem. We decided to use affinity propagation clustering.[^45] That algorithm assumes that each cluster has an "exemplar" item which the other items are similar to. That assumption fits nicely with borrowings from the Field Code, where a single section (likely from the 1850 New York code) had many borrowings, but where there could also be innovative sections in other stages.[^46]

The result was a set of approximately 2,900 clusters which contained at least five sections, though this probably overstates the number of innovative, *ur*-sections in the corpus. Each cluster contained a list of the section that belonged to it. The biggest cluster, for instance, which concerned the use of affidavits in pleading, contained 103 sections. With those clusters, we were able to organize the documents chronologically.[^47] This method is particularly useful when tracking the variations in in the law from jurisdiction to jurisdiction. It provides historians with a way of noticing small changes in the wording and substance of the law. Most discussions of algorithmic reading have focused on "distant reading,"[^48] or have balanced the claims of distant reading by using it as a means to enable close reading. This method of clustering, however, is a kind of algorithmic close reading. By deforming the texts---taking them out of the context of the codes and putting them in the context of their particular variations---we are able to pay attention to those variations.[^49]

Take for instance, provisions regulating witness testimony. At common law, parties and interested witnesses were not permitted to testify in their own causes. Field's Code reversed this rule, expanding witness competency as widely as possible: any person "having organs of sense" could now be admitted as a witness in New York, with only the insane and very young children exempted from the rule. As the Code migrated West, however, legislators added racial exclusions to Field's list. The following cluster documents how California's codifiers engrafted earlier prohibitions from midwestern states into Field's Code. Many other codes then evidence a remarkable uniformity with California's text (which later changed only to add "Mongolian" to the list of races). Iowa's Code had much more minor influence, and Wyoming developed the only truly original section which made clear the link to a racist assumption of infantilization of non-white peoples. \[**Figure: Cluster from Kellen**\]

As these clusters illustrate, uniformity can be just as instructive as variation. On the point, the most significant clusters that we investigated related to the collection of debts. These were clusters which went against the typical pattern that we observed. While many clusters exhibited regional variation as they grew more distant from the Field Code, clusters having to do with creditors' remedies were almost completely uniform across the American West and South. No single section of the Code announced its preference for creditors' rights; rather, the acceleration of creditors' remedies resulted from the combination of several sections. In the original enacted Code, § 107 required a defendant to answer the complaint within twenty days, instead of at the next court session (which in some cases could have been as far as three months away); § 202 provided for default judgment as a matter of course, issued by a clerk without a judicial order if no adequate answer was received within the twenty days; §§ 128-133 abolished fictitious pleadings and required answers to state true facts verified by a defendant's oath, all so that no trial would delay the enforcement of uncontestable obligations; finally, the Code abolished a traditional thirty-day waiting period between issue of judgment and commencement of execution (see First Report (New York 1848), 197). These provisions dealt with what merchants and capitalists perceived as an abuse of the common law system, where defendants in cases of debt could stretch out enforcement of debt collection for as long as two years. The Field code's summary judgment brought down the time of debt collection to a matter of weeks. Effectively, the Code traded the rhythms of agriculture for the rhythms of merchant finance.[^50]

Clustering each of these sections reveals that western states along with the former Confederate states of South Carolina, North Carolina, and Florida copied each provision almost exactly. Midwestern and Upper South states that had already developed by 1850 and maintained commercial ties to Chicago and New Orleans varied the New York rules, sometimes by requiring answers only in term time, or permitting only a judge to decree default judgment rather than a clerk, in either case effectively stretching out enforcement and making a formal trial more likely. But in the Reconstruction South, and in the West over the same period, regardless of whether a jurisdiction abolished chancery or not, regardless of the racial exclusions it may have placed on witness testimony, the provisions on debt collection remained unchanged. When it came to creditors' remedies, the law of New York became the law of the land.

**Part 4**.

Alerted by these clustering patterns to the significance of creditors' remedies in the developing West and re-developing South, it was not difficult to find evidence of westerners and southerners commenting on these features of the Code or its connections to New York capital. Twelve of the states and territories that copied the Field Code mostly closely did so during the Civil War and Reconstruction era---four states in the former Confederacy and eight jurisdictions in the Far West.[^51] As with other areas of postbellum historiography, it turns out one may learn a lot by holding the postellum American South and American West together.

In the last decade, scholars of Reconstruction have broadened the scope of their study to include both the South and the West as two sites in one "Greater Reconstruction." These studies have illustrated the ways that military conquest, rapid industrialization, and the resettlement and education of ethnic minorities developed similarly in each region, guided by economic elites in New York and political administrators in Washington.[^52] In tracing the legal aspects of this Greater Reconstruction, scholars have focused almost entirely on constitutional rights of citizenship and civil equality or the expansion of federal power.[^53] While the 1860s and 1870s were of course a transformative period in the history of civil rights and the creation of a national state, they were also the decades in which many local legal institutions and practices were transformed not by federal power but by state codification. Naomi Lamoreaux and John Joseph Wallis have recently argued that in the creation of a modern American economy, "the federal government played *no* role in this process until the Civil War, and even then it played only a bit part." The history of the Field Code's migration helps to substantiate this claim. While Lamoreaux and Wallis focus on the development of banking, transportation, and incorporation at the state level, it was the Procedure Code that provided the structure of civil remedies that protected these state institutions. And procedure codes were, of course, the creatures of state governance.[^54]

With the universal complaints that time was too short or commissioners too democratically irresponsible, one might expect the Code to have faced more competition from other workable statutes. Southern states that drew more on the civilian code tradition such as Louisiana and Alabama offered other options for states desiring to reform common law practice.[^55] And if Euro-Americans believed law had to be imported into the West, Colorado immigrants proved that the common law of Illinois was just as amenable to statutory importation as New York law.[^56] But what codifiers saw when they looked at New York, more so than Louisiana, Alabama, or Illinois, was the Empire State of commercial capital. And the putative fears, demands, and desires of a personified Capital continually wielded promises---and threats---in the debates over procedural codification. The early Mormon settlers of Utah persistently avoided the mining frenzy as well as codification. By 1870, however, the territory's governor directed the legislature's attention to the recent Code of Nevada, a code "for a people whose interests in many respects are similar to our own." Of course, standing behind the Nevada Code was "the State of New York---a State which is an empire in itself and whose commercial transactions are far greater than those of any other State in the Union." By copying its code, Utah too could be "rewarded by equal advantages."[^57] Code proponents in Colorado similarly parried arguments that a code would be a "foreign" imposition by pointing to the fact that it has been "adopted twenty-nine years ago by the Empire state of the Union" and they promised that the code of the nation's uniquely commercial empire could bring wealth in its wake.[^58]

When a Colorado legislator scoffed at the idea that capitalists could possibly care about the difference between old common law and modern code remedies, his adversaries rebuked him. "Mr. Hamill replied that he knew of one California company of capitalists who were deterred from investing in mining property here wholly on account of the practice of the courts in mining cases. If we had had this code years ago, Colorado would now have a larger amount of California capital in her mines."[^59] Codifiers argued that, in attracting capital, procedure was at least as important as the substantive rules of property and contract, because procedure secured the remedies that actually protected investments. "Men of capital and enterprise will not make investments and devote their time and energies to those works of internal improvement so necessary for the speedy development of our natural resources, and for the settlement and building up the country," Nebraska's governor reasoned, "unless ample protection is afforded them, by legal enactment, for the capital invested and labor employed." And he therefore urged swift passage of the Field Code.[^60]

Receiving innumerable letters complaining that under the Code, "no one will be benefited, except perhaps some Northern Capitalists," a North Carolina commissioner undertook an anonymous defense of the new Code in the *Weekly Standard*. He encouraged the bar to accommodate themselves to change, for "the New York system . . . bids fair to become national." Putatively offering an overview of the Code, the articles were almost entirely about credit. "How can we create credit? By punctuality," the commissioner wrote. "And how create punctuality? by law, and by law alone. Let the law enforce punctuality; let the people of North Carolina learn that the great law of business is, that 'time is of the essence of the contract,' and incur no debt that they do not expect to meet at maturity." Under the more certain and speedy remedies of the Code, "we may expect that the secret hoards of the frugal among ourselves will be offered to loan, and even that the vaults of the banks of New-York . . . will be open to our industry."[^61]

As in postbellum North Carolina, establishing a flow of credit through the remedial system became a leading priority of western lawyers. While the so-called new western history has shed significant light on neglected topics of Indian dispossession and environmental management, it has often done so by leaving out of view matters of political economy, a staple of the old western history. As one work in that older tradition argued, "Debt collection was the central part of law practice for the \[western\] bar and remained a key part of private practice throughout the century."[^62] On that understanding, one western lawyer succinctly summarized the difference between the Code and the common law as "whether a merchant had better try to collect a \$500 note or burn it up." Tiring of all the focus on creditors' remedies, one legislator observed that he "never knew one of these professionals who undertook to write up the beauties of the New York code, . . . that he did not also break out somewhere with 'take for instance the case of an action on a promissory note,' as though the collection of notes was about all there could be any law needed for."[^63]

The creditors' remedies in the Code gave the codifiers their leading argument against criticisms rooted in the ideology of popular sovereignty. "There is no doubt but the people are in favor of anything that promises to hurry up that proverbially slow and blind old female called Justice, and they will go for the old code," one Colorado newspaper announced.[^64] New York's "code practice is the best in excellence," stated another, "and when I say *best* I do not mean best for lawyers only, but best for the people---the commonwealth."[^65] If the People favored economic progress, certainty of remedy, and efficiency in proceedings, then they favored the New York Code, no matter whether they understood or cared about technical rules of pleading and joinder. Thus, in their arguments the codifiers imagined themselves the champions of popular sovereignty, for it was they who accomplished what the People actually desired.
